Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/semantics.tex
tweak semantics

The witness generation procedure @genWitness@ is formalized in
Figure~\ref{fig:algo-gen-witness}.
%
Next, we describe what it takes as input and output, and how it
addresses the above challenges to search the space of possible
executions for general type error witnesses.

\paragraph{Inputs and Outputs}
%
Obviously, the problem of generating inputs is undecidable in general.
%
The witness generation procedure takes two inputs.
%
First, a search bound @n@ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
denoting the list of \emph{top-level definitions}
%
\begin{code}
  let f1 = e1 in
  let f2 = e2 in
  ...
  let fn = en in
\end{code}
%
The witness generation procedure returns as output a list of (general)
witness expressions $e$ each of which is of the form @fn v1 ... vn@.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.
%
In the sequel, we refer to @fn@ as the \emph{target} of witness generation.

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\begin{mcode}
  eval :: ($e$, $\su$) -> [($v$, $\su$)]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call \hbox{|eval($e$, $\vsu$)|}
returns the \emph{list} possible results of the form $\pair{e'}{\vsu'}$ such
that $\steps{e}{\vsu}{e'}{\vsu'}$.
%
We determine whether a given value is stuck by using the predicate
%
\begin{mcode}
  isStuck :: $v$ -> Bool
\end{mcode}

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that \\
\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes until
@eval@uating the closed term yields a non-lambda value.
%
The helper functions @witness@ and @close@ are used to respectively
apply the parameters to the target and close the result under the top-level
let-binders, prior to invoking @eval@.
%
That is, @mkApps@ creates a nested sequence of applications in
the usual left-associative style, and @mkLets@ takes a list of
binders and a body expression, and creates a sequence of nested
let-binders that close the body expression.

@@

\begin{thm}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k@ $[(f_1, e_1), \ldots, (f_n, e_n)]$ is non-empty,
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}

>>>>

The witness generation procedure @genWitness@ is formalized in
Figure~\ref{fig:algo-gen-witness}.
%
Next, we describe what its input and output, and how it
addresses the above challenges to search the space of possible
executions for general type error witnesses.

\paragraph{Inputs and Outputs}
%
Obviously, the problem of generating inputs is undecidable in general.
%
The witness generation procedure takes two inputs.
%
First, a search bound @n@ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, binder-expression tuples @[(f1,e1),...,(fn,en)]@
denoting the list of \emph{top-level definitions}:
%
\begin{code}
  let f1 = e1 in
  let f2 = e2 in
  ...
  let fn = en in
\end{code}
%
The witness generation procedure returns as output a list of (general)
witness expressions $e$ each of which is of the form @fn v1 ... vn@.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.
%
In the sequel, we refer to @fn@ as the \emph{target} of witness generation.

\paragraph{Modeling Semantics}
%: \texttt{eval} and \texttt{isStuck}}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure:
%
\begin{mcode}
  eval :: ($e$, $\su$) -> [($v$, $\su$)]
\end{mcode}
%
Due to the non-determinism introduced by \gensym, a call \hbox{|eval($e$, $\vsu$)|}
returns the \emph{list} possible results of the form $\pair{e'}{\vsu'}$ such
that $\steps{e}{\vsu}{e'}{\vsu'}$.
%
We determine whether a given value is stuck by using the predicate:
%
\begin{mcode}
  isStuck :: $v$ -> Bool
\end{mcode}

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that \\
\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes until
@eval@uating the closed term yields a non-lambda value.
%
The helper functions @witness@ and @close@ are used to respectively
apply the parameters to the target and close the result under the top-level
let-binders, prior to invoking \hbox{@eval@.}
%
That is, @mkApps@ creates a nested sequence of applications in
the usual left-associative style, and @mkLets@ takes a list of
binders and a body expression, and creates a sequence of nested
let-binders that close the body expression.

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\stuck}{\su}$.
\end{cor}
